Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-03-15 paper/ms.tex
typos and edits

In the case of $J$ \celerite\ terms, we can check for negative values of the
PSD by solving for the roots of the power spectrum, abbreviating with
$z = \omega^2$:
\begin{equation}
P(\omega)=  \sum_{j=1}^J \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
\end{equation}
where
\begin{eqnarray}
q_j &=& a_jc_j-b_jd_j\\
r_j &=& (d_j^2+c_j^2)(b_jd_j+a_jc_j)\\
s_j &=& 2(c_j^2-d_j^2)\\
t_j &=& (c_j^2+d_j^2)^2.
\end{eqnarray}
The denominators of each term are positive, so we can multiply through by $\Pi_j \left(z^2+s_jz + t_j\right)$ yielding:
\begin{equation}
P_0(z) = \sum_{j=1}^J (q_j z + r_j)\Pi_{k \ne j}\left(z^2+s_kz + t_k\right) = 0,
\end{equation}
which is a polynomial with order $p_{ord}=2(J-1)+1$.  With $J=2$, this yields a cubic equation
which may be solved exactly for the roots.

For arbitrary $J$, a procedure based upon Sturm's theorem \citep{Dorrie:1965} allows one to determine whether there are any
real roots within the range $(0,\infty]$. We first construct $P_0(z)$ and it's derivative
$P_1(z) = P^\prime(z)$, and then loop from $k=2$ to $k=p_{ord}$, computing
$P_k(z) = -{\rm rem}(P_{k-2},P_{k-1})$.  The function ${\rm rem}(p,q)$ is the remainder polynomial after dividing $p(z)$ by $q(z)$.

We evaluate the $z^0$ coefficients of each of the polynomial in the series by evaluating $f_0 = \{P_0(0),...,P_{p_{ord}}(0)\}$ to
give us the signs of these polynomials evaluated at $z=0$.
Likewise, we evaluate the coefficients of the largest order term in each polynomial which gives the sign of the polynomial
as $z \rightarrow \infty$.  This gives $f_\infty = \{C(P_0,p_{ord}),C(P_1,p_{ord}-1),..., C(P_{p_{ord}},1)\}$ where
$C(p(z),m)$ returns the coefficient of $z^m$ in polynomial $p(z)$.

With the series of coefficients $f_0$ and $f_\infty$, we then determine how many times the sign changes in each
of these, where $\sigma(0)$ is the number of sign changes at $z=0$, and $\sigma(\infty)$ is the number of sign
changes at $z \rightarrow \infty$.  The total number of real roots in the range $(0,\infty]$ is given
by $N_{+}=\sigma(0)-\sigma(\infty)$.

We have checked that this procedure works for a wide range of parameters, and we find that it robustly
matches the number of positive real roots which we evaluated numerically.

The advantage of this procedure is that it does not require computing the roots, but only carrying out algebraic
manipulation of polynomials to determine the number of positive real roots.  If a non-zero real root is found, then
likelihood may be set to zero.

We have implemented this algorithm in subroutines that may be called to check for positive-definiteness
of a kernel with some values of $a_j$ or $b_j$ negative.
%Sturm's theorem applied to numerator of power spectrum.

@@

>>>>

@@

In the case of $J$ \celerite\ terms, we can check for negative values of the
PSD by solving for the roots of the power spectrum (see \eqalt{celerite-psd}),
abbreviating with $z = \omega^2$:
\begin{equation}
S(\omega)=  \sum_{j=1}^J \frac{q_j z + r_j}{z^2+s_jz + t_j} = 0
\end{equation}
where
\begin{eqnarray}
q_j &=& a_jc_j-b_jd_j\\
r_j &=& (d_j^2+c_j^2)(b_jd_j+a_jc_j)\\
s_j &=& 2(c_j^2-d_j^2)\\
t_j &=& (c_j^2+d_j^2)^2.
\end{eqnarray}
The denominators of each term are positive, so we can multiply through by
$\prod_j \left(z^2+s_jz + t_j\right)$ to find
\begin{equation}
Q_0(z) = \sum_{j=1}^J (q_j z + r_j)\prod_{k \ne j}\left(z^2+s_kz +
    t_k\right) = 0\quad,
\end{equation}
which is a polynomial with order $2\,(J-1)+1$.
With $J=2$, this yields a cubic equation which may be solved exactly for the
roots.

For arbitrary $J$, a procedure based upon Sturm's theorem \citep{Dorrie:1965}
allows one to determine whether there are any real roots within the range
$(0,\,\infty]$.
We first construct $Q_0(z)$ and it's derivative $Q_1(z) = {Q_0}^\prime(z)$,
and then loop from $k=2$ to $k=2\,(J-1)+1$, computing
\begin{eqnarray}
Q_k(z) = -{\rm rem}(Q_{k-2},\,Q_{k-1})
\end{eqnarray}
where the function ${\rm rem}(p,\,q)$ is the remainder polynomial after
dividing $p(z)$ by $q(z)$.

We evaluate the coefficients of each of the polynomial in the series by
evaluating $f_0 = \{Q_0(0),\,\ldots,\,Q_{2\,(J-1)+1}(0)\}$ to give us the signs
of these polynomials evaluated at $z=0$.
Likewise, we evaluate the coefficients of the largest order term in each
polynomial which gives the sign of the polynomial as $z \rightarrow \infty$,
$f_\infty$.
With the series of coefficients $f_0$ and $f_\infty$, we then determine how
many times the sign changes in each of these, where $\sigma(0)$ is the number
of sign changes at $z=0$, and $\sigma(\infty)$ is the number of sign changes
at $z \rightarrow \infty$.
The total number of real roots in the range
$(0,\,\infty]$ is given by $N_{+}=\sigma(0)-\sigma(\infty)$.

We have checked that this procedure works for a wide range of parameters, and
we find that it robustly matches the number of positive real roots which we
evaluated numerically.
The advantage of this procedure is that it does not require computing the
roots, but only carrying out algebraic manipulation of polynomials to
determine the number of positive real roots.
If a non-zero real root is found, the likelihood may be set to zero.
